---
title: "creating_infiles, rlpi package"
author: "Robin Freeman, IoZ, Zoological Society of London"
date: "`r format(Sys.Date())`"
output:  
      html_document:  
        keep_md: true  
        toc: true
        theme: united
---

```{r, echo=FALSE}
require(rlpi)
```

The code also provides a means to create infiles from tabular population data. For example, the Living Planet Database can be constructed so that each row represents a population, with columns for each abundance year. NB: The **create_infile** method expects particular columns that define where the abundance data resides, it uses the **convert_to_rows** function that assumes abundance data is in columns between the *X1950* column and a column called *Managed*. This reflects the format used in the Living Planet Database, and in the example data provided here **http://www.livingplanetindex.org/projects?main_page_project=LivingPlanetReport&home_flag=1**.


```{r making_infiles, eval=T}
# Constructing infiles from a populations table...

# First read the population table (this is the Living Planet Database excluding confidential records)
lpi_data <- read.csv("inst/extdata/example_data/LPI_LPR2016data_public.csv", na.strings = "NULL")

# Create an inifle from all the data. All the population data in the 'lpi_data' table will be converted and stored in a file called 'example_data_pops.txt' and a file called 'example_data_infile.txt' will be created that references the first file (the infile name will also be stored in the returned variable 'example_infile_name')

# Here we select the first 100 populations by creating an index vector that's FALSE for all rows, then setting the first 100 rows to TRUE
index_vector = rep(FALSE, nrow(lpi_data))
index_vector[1:100] = TRUE

example_infile_name <- create_infile(lpi_data, index_vector=index_vector, name="example_data")
# An index can be created using this infile, for the period 1970 to 2014 with 100 bootstraps.
example_lpi <- LPIMain(example_infile_name, REF_YEAR = 1970, PLOT_MAX = 2014, BOOT_STRAP_SIZE = 100)

# Plot the resulting index
ggplot_lpi(example_lpi, title = "example_lpi", xlims=c(1970, 2012), ylim=c(0, 2))
# Plot the resulting index with logged y-axis (note the use of a +ive ymin)
ggplot_lpi(example_lpi, title = "example_lpi", xlims=c(1970, 2012), ylim=c(0.3, 2), trans="log")

# Here we limit the data to 'Strigiformes' simply by creating a boolean  (true/false) vector which is true for populations (rows) where the Order is "Strigiformes"
Strigiformes = lpi_data$Order == "Strigiformes" 

# Psssing this vector into the create_infile function will select just those populations and create an infile for them
s_infile_name <- create_infile(lpi_data, index_vector=Strigiformes, name="example_data_strig")
# Again, create and index
s_lpi <- LPIMain(s_infile_name, REF_YEAR = 1970, PLOT_MAX = 2014, BOOT_STRAP_SIZE = 100)
# And plot
ggplot_lpi(s_lpi, title = "s_lpi", xlims=c(1970, 2012))

# Similarly, this will create an index just for those populations of the Order 'Passeriformes'
Passeriformes = lpi_data$Order == "Passeriformes" 
p_infile_name <- create_infile(lpi_data, index_vector=Passeriformes, name="example_data_pass")
p_lpi <- LPIMain(s_infile_name, REF_YEAR = 1970, PLOT_MAX = 2014, BOOT_STRAP_SIZE = 100)
ggplot_lpi(p_lpi, title = "p_lpi", xlims=c(1970, 2012))

# Nearctic mammals
Nearctic_mammals = lpi_data$Class == "Mammalia" & lpi_data$T_realm == "Nearctic"
nm_infile_name <- create_infile(lpi_data, index_vector=Nearctic_mammals, name="terrestrial_Nearctic_Mammalia")

# How many pops...
sum(Nearctic_mammals, na.rm = T)

nm_lpi <- LPIMain(nm_infile_name, REF_YEAR = 1970, PLOT_MAX = 2014, BOOT_STRAP_SIZE = 100)
ggplot_lpi(nm_lpi, title = "nm_lpi", xlims=c(1970, 2012))


# Nearctic birds
Nearctic_birds = lpi_data$Class == "Aves" & lpi_data$T_realm == "Nearctic"
nb_infile_name <- create_infile(lpi_data, index_vector=Nearctic_birds, name="terrestrial_Nearctic_Aves")

# How many pops...
sum(Nearctic_birds, na.rm = T)

nb_lpi <- LPIMain(nb_infile_name, REF_YEAR = 1970, PLOT_MAX = 2014, BOOT_STRAP_SIZE = 100)
ggplot_lpi(nb_lpi, title = "nb_lpi", xlims=c(1970, 2012))


# Nearctic herps
Nearctic_herps = (lpi_data$Class == "Reptilia" | lpi_data$Class == "Amphibia") & lpi_data$T_realm == "Nearctic"
nh_infile_name <- create_infile(lpi_data, index_vector=Nearctic_herps, name="terrestrial_Nearctic_Herps")

# How many pops...
sum(Nearctic_herps, na.rm = T)

nh_lpi <- LPIMain(nh_infile_name, REF_YEAR = 1970, PLOT_MAX = 2014, BOOT_STRAP_SIZE = 100)
ggplot_lpi(nh_lpi, title = "nh_lpi", xlims=c(1970, 2012))

```

```{r making_infiles, eval=T}
selected_pops = lpi_data$Class == "Aves" & lpi_data$T_realm == "Afrotropical"
sum(selected_pops, na.rm = T)
infile_name <- create_infile(lpi_data, index_vector=selected_pops, name="T_Afrotropical_aves")

selected_pops = lpi_data$Class == "Aves" & (lpi_data$T_realm == "Indo-Malayan" | lpi_data$T_realm == "Australasia"  | lpi_data$T_realm == "Oceania")
sum(selected_pops, na.rm = T)
infile_name <- create_infile(lpi_data, index_vector=selected_pops, name="T_IndoPacific_aves")

selected_pops = lpi_data$Class == "Aves" & lpi_data$T_realm == "Palearctic"
sum(selected_pops, na.rm = T)
infile_name <- create_infile(lpi_data, index_vector=selected_pops, name="T_Palearctic_aves")

selected_pops = lpi_data$Class == "Aves" & lpi_data$T_realm == "Neotropical"
sum(selected_pops, na.rm = T)
infile_name <- create_infile(lpi_data, index_vector=selected_pops, name="T_Neotropical_aves")

selected_pops = lpi_data$Class == "Aves" & lpi_data$T_realm == "Nearctic"
sum(selected_pops, na.rm = T)
infile_name <- create_infile(lpi_data, index_vector=selected_pops, name="T_Nearctic_aves")



selected_pops = lpi_data$Class == "Mammalia" & lpi_data$T_realm == "Afrotropical"
sum(selected_pops, na.rm = T)
infile_name <- create_infile(lpi_data, index_vector=selected_pops, name="T_Afrotropical_mammalia")

selected_pops = lpi_data$Class == "Mammalia" & (lpi_data$T_realm == "Indo-Malayan" | lpi_data$T_realm == "Australasia"  | lpi_data$T_realm == "Oceania")
sum(selected_pops, na.rm = T)
infile_name <- create_infile(lpi_data, index_vector=selected_pops, name="T_IndoPacific_mammalia")

selected_pops = lpi_data$Class == "Mammalia" & lpi_data$T_realm == "Palearctic"
sum(selected_pops, na.rm = T)
infile_name <- create_infile(lpi_data, index_vector=selected_pops, name="T_Palearctic_mammalia")

selected_pops = lpi_data$Class == "Mammalia" & lpi_data$T_realm == "Neotropical"
sum(selected_pops, na.rm = T)
infile_name <- create_infile(lpi_data, index_vector=selected_pops, name="T_Neotropical_mammalia")

selected_pops = lpi_data$Class == "Mammalia" & lpi_data$T_realm == "Nearctic"
sum(selected_pops, na.rm = T)
infile_name <- create_infile(lpi_data, index_vector=selected_pops, name="T_Nearctic_mammalia")



selected_pops = (lpi_data$Class == "Reptilia" | lpi_data$Class == "Amphibia") & lpi_data$T_realm == "Afrotropical"
sum(selected_pops, na.rm = T)
infile_name <- create_infile(lpi_data, index_vector=selected_pops, name="T_Afrotropical_herps")

selected_pops = (lpi_data$Class == "Reptilia" | lpi_data$Class == "Amphibia") & (lpi_data$T_realm == "Indo-Malayan" | lpi_data$T_realm == "Australasia"  | lpi_data$T_realm == "Oceania")
sum(selected_pops, na.rm = T)
infile_name <- create_infile(lpi_data, index_vector=selected_pops, name="T_IndoPacific_herps")

selected_pops = (lpi_data$Class == "Reptilia" | lpi_data$Class == "Amphibia") & lpi_data$T_realm == "Palearctic"
sum(selected_pops, na.rm = T)
infile_name <- create_infile(lpi_data, index_vector=selected_pops, name="T_Palearctic_herps")

selected_pops = (lpi_data$Class == "Reptilia" | lpi_data$Class == "Amphibia") & lpi_data$T_realm == "Neotropical"
sum(selected_pops, na.rm = T)
infile_name <- create_infile(lpi_data, index_vector=selected_pops, name="T_Neotropical_herps")

selected_pops = (lpi_data$Class == "Reptilia" | lpi_data$Class == "Amphibia") & lpi_data$T_realm == "Nearctic"
sum(selected_pops, na.rm = T)
infile_name <- create_infile(lpi_data, index_vector=selected_pops, name="T_Nearctic_herps")


T_Afrotropical_aves_pops.txt	1	0.387205957	0.189737662
T_Afrotropical_mammalia_pops.txt	1	0.197833813	0.189737662
T_Afrotropical_herps_pops.txt	1	0.41496023	0.189737662

T_IndoPacific_aves_pops.txt	2	0.396527091	0.292168385
T_IndoPacific_mammalia_pops.txt	2	0.172106825	0.292168385
T_IndoPacific_herps_pops.txt	2	0.431366084	0.292168385

T_Palearctic_aves_pops.txt	3	0.433535576	0.116430659
T_Palearctic_mammalia_pops.txt	3	0.249862107	0.116430659
T_Palearctic_herps_pops.txt	3	0.316602317	0.116430659

T_Neotropical_aves_pops.txt	4	0.387661234	0.321131554
T_Neotropical_mammalia_pops.txt	4	0.127987201	0.321131554
T_Neotropical_herps_pops.txt	4	0.484351565	0.321131554

T_Nearctic_aves_pops.txt	5	0.376366476	0.061683203
T_Nearctic_mammalia_pops.txt	5	0.249869859	0.061683203
T_Nearctic_herps_pops.txt	5	0.373763665	0.061683203

```
