---
title: "creating_infiles, rlpi package"
author: "Louise McRae, Stefanie Deinet, Robin Freeman, IoZ, Zoological Society of London"
output:  
      html_document:  
        keep_md: true  
        toc: true
        theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```

```{r, echo=FALSE, message=FALSE}
require(rlpi)
```

The code also provides a means to create infiles from tabular population data such as provided  in the comma separated output of Living Planet Database **[here](https://dx.doi.org/10.6084/m9.figshare.4300022.v1)**. NB: The **create_infile** method currently expects particular columns that define where the abundance data resides, using the **convert_to_rows** function that assumes abundance data is in columns between the *X1950* column and a column called *Managed*. 

> Note: This [Comma Separated Value version of the Living Planet Database ](https://dx.doi.org/10.6084/m9.figshare.4300022.v1) excludes around 3000 populations which are confidential and cannot therefore be shared. Therefore, results produced with this data set may differ slightly from those presented in the manuscript and elsewhere.

```{r rlpi_data, eval=FALSE}
# Get example data from package
# Copy zipped data to local directory 
file.copy(from=system.file("extdata", "example_data.zip", package = "rlpi"), to=getwd())
# Extract data, this will create a directory of terrestrial LPI data to construct a terrestrial index from.
unzip("example_data.zip")

```

```{r making_infiles, eval=TRUE, message=FALSE}
# Constructing infiles from a populations table...

# First read the population table (this is the Living Planet Database excluding confidential records)
lpi_data <- read.csv("example_data/LPI_LPR2016data_public.csv", na.strings = "NULL")

# Create an infile from all the data. All the population data in the 'lpi_data' table will be converted and stored in a file called 'example_data_pops.txt' and a file called 'example_data_infile.txt' will be created that references the first file (the infile name will also be stored in the returned variable 'example_infile_name')

# Here we select the first 100 populations by creating an index vector that's FALSE for all rows, then setting the first 100 rows to TRUE
index_vector = rep(FALSE, nrow(lpi_data))
index_vector[1:100] = TRUE

example_infile_name <- create_infile(lpi_data, index_vector=index_vector, name="example_data")
# An index can be created using this infile, for the period 1970 to 2014 with 100 bootstraps.
example_lpi <- LPIMain(example_infile_name, REF_YEAR = 1970, PLOT_MAX = 2014, BOOT_STRAP_SIZE = 100, VERBOSE=FALSE, show_progress=FALSE)

# Remove NAs (trailing years with no data)
example_lpi <- example_lpi[complete.cases(example_lpi), ]

# Plot the resulting index
ggplot_lpi(example_lpi, title = "example_lpi", xlims=c(1970, 2012), ylim=c(0, 2))
# Plot the resulting index with logged y-axis (note the use of a +ive ymin)
ggplot_lpi(example_lpi, title = "example_lpi", xlims=c(1970, 2012), ylim=c(0.3, 2), trans="log")

# Here we limit the data to 'Strigiformes' simply by creating a boolean  (true/false) vector which is true for populations (rows) where the Order is "Strigiformes"
Strigiformes = lpi_data$Order == "Strigiformes" 

# Passing this vector into the create_infile function will select just those populations and create an infile for them
s_infile_name <- create_infile(lpi_data, index_vector=Strigiformes, name="example_data_strig")
# Again, create and index
s_lpi <- LPIMain(s_infile_name, REF_YEAR = 1970, PLOT_MAX = 2014, BOOT_STRAP_SIZE = 100, VERBOSE=FALSE, show_progress=FALSE)
# And plot

# Remove NAs (trailing years with no data)
s_lpi <- s_lpi[complete.cases(s_lpi), ]
ggplot_lpi(s_lpi, title = "s_lpi", xlims=c(1970, 2012))

# Similarly, this will create an index just for those populations of the Order 'Passeriformes'
Passeriformes = lpi_data$Order == "Passeriformes" 
p_infile_name <- create_infile(lpi_data, index_vector=Passeriformes, name="example_data_pass")
p_lpi <- LPIMain(s_infile_name, REF_YEAR = 1970, PLOT_MAX = 2014, BOOT_STRAP_SIZE = 100, VERBOSE=FALSE, show_progress=FALSE)

# Remove NAs (trailing years with no data)
p_lpi <- p_lpi[complete.cases(p_lpi), ]
ggplot_lpi(p_lpi, title = "p_lpi", xlims=c(1970, 2012))

# Nearctic mammals
Nearctic_mammals = lpi_data$Class == "Mammalia" & lpi_data$T_realm == "Nearctic"
nm_infile_name <- create_infile(lpi_data, index_vector=Nearctic_mammals, name="terrestrial_Nearctic_Mammalia")

# How many pops...
sum(Nearctic_mammals, na.rm = T)

nm_lpi <- LPIMain(nm_infile_name, REF_YEAR = 1970, PLOT_MAX = 2014, BOOT_STRAP_SIZE = 100, VERBOSE=FALSE, show_progress=FALSE)

# Remove NAs (trailing years with no data)
nm_lpi <- nm_lpi[complete.cases(nm_lpi), ]
ggplot_lpi(nm_lpi, title = "nm_lpi", xlims=c(1970, 2012))


# Nearctic birds
Nearctic_birds = lpi_data$Class == "Aves" & lpi_data$T_realm == "Nearctic"
nb_infile_name <- create_infile(lpi_data, index_vector=Nearctic_birds, name="terrestrial_Nearctic_Aves")

# How many pops...
sum(Nearctic_birds, na.rm = T)

nb_lpi <- LPIMain(nb_infile_name, REF_YEAR = 1970, PLOT_MAX = 2014, BOOT_STRAP_SIZE = 100, VERBOSE=FALSE, show_progress=FALSE)

# Remove NAs (trailing years with no data)
nb_lpi <- nb_lpi[complete.cases(nb_lpi), ]
ggplot_lpi(nb_lpi, title = "nb_lpi", xlims=c(1970, 2012))


# Nearctic herps
Nearctic_herps = (lpi_data$Class == "Reptilia" | lpi_data$Class == "Amphibia") & lpi_data$T_realm == "Nearctic"
nh_infile_name <- create_infile(lpi_data, index_vector=Nearctic_herps, name="terrestrial_Nearctic_Herps")

# How many pops...
sum(Nearctic_herps, na.rm = T)

nh_lpi <- LPIMain(nh_infile_name, REF_YEAR = 1970, PLOT_MAX = 2014, BOOT_STRAP_SIZE = 100, VERBOSE=FALSE, show_progress=FALSE)

# Remove NAs (trailing years with no data)
nh_lpi <- nh_lpi[complete.cases(nh_lpi), ]
ggplot_lpi(nh_lpi, title = "nh_lpi", xlims=c(1970, 2012))

```

Constructing infiles

```{r making_infiles_for_terr, echo=FALSE, eval=FALSE}
selected_pops = lpi_data$Class == "Aves" & lpi_data$T_realm == "Afrotropical"
sum(selected_pops, na.rm = T)
infile_name <- create_infile(lpi_data, index_vector=selected_pops, name="T_Afrotropical_aves")

selected_pops = lpi_data$Class == "Aves" & (lpi_data$T_realm == "Indo-Malayan" | lpi_data$T_realm == "Australasia"  | lpi_data$T_realm == "Oceania")
sum(selected_pops, na.rm = T)
infile_name <- create_infile(lpi_data, index_vector=selected_pops, name="T_IndoPacific_aves")

selected_pops = lpi_data$Class == "Aves" & lpi_data$T_realm == "Palearctic"
sum(selected_pops, na.rm = T)
infile_name <- create_infile(lpi_data, index_vector=selected_pops, name="T_Palearctic_aves")

selected_pops = lpi_data$Class == "Aves" & lpi_data$T_realm == "Neotropical"
sum(selected_pops, na.rm = T)
infile_name <- create_infile(lpi_data, index_vector=selected_pops, name="T_Neotropical_aves")

selected_pops = lpi_data$Class == "Aves" & lpi_data$T_realm == "Nearctic"
sum(selected_pops, na.rm = T)
infile_name <- create_infile(lpi_data, index_vector=selected_pops, name="T_Nearctic_aves")



selected_pops = lpi_data$Class == "Mammalia" & lpi_data$T_realm == "Afrotropical"
sum(selected_pops, na.rm = T)
infile_name <- create_infile(lpi_data, index_vector=selected_pops, name="T_Afrotropical_mammalia")

selected_pops = lpi_data$Class == "Mammalia" & (lpi_data$T_realm == "Indo-Malayan" | lpi_data$T_realm == "Australasia"  | lpi_data$T_realm == "Oceania")
sum(selected_pops, na.rm = T)
infile_name <- create_infile(lpi_data, index_vector=selected_pops, name="T_IndoPacific_mammalia")

selected_pops = lpi_data$Class == "Mammalia" & lpi_data$T_realm == "Palearctic"
sum(selected_pops, na.rm = T)
infile_name <- create_infile(lpi_data, index_vector=selected_pops, name="T_Palearctic_mammalia")

selected_pops = lpi_data$Class == "Mammalia" & lpi_data$T_realm == "Neotropical"
sum(selected_pops, na.rm = T)
infile_name <- create_infile(lpi_data, index_vector=selected_pops, name="T_Neotropical_mammalia")

selected_pops = lpi_data$Class == "Mammalia" & lpi_data$T_realm == "Nearctic"
sum(selected_pops, na.rm = T)
infile_name <- create_infile(lpi_data, index_vector=selected_pops, name="T_Nearctic_mammalia")

selected_pops = (lpi_data$Class == "Reptilia" | lpi_data$Class == "Amphibia") & lpi_data$T_realm == "Afrotropical"
sum(selected_pops, na.rm = T)
infile_name <- create_infile(lpi_data, index_vector=selected_pops, name="T_Afrotropical_herps")

selected_pops = (lpi_data$Class == "Reptilia" | lpi_data$Class == "Amphibia") & (lpi_data$T_realm == "Indo-Malayan" | lpi_data$T_realm == "Australasia"  | lpi_data$T_realm == "Oceania")
sum(selected_pops, na.rm = T)
infile_name <- create_infile(lpi_data, index_vector=selected_pops, name="T_IndoPacific_herps")

selected_pops = (lpi_data$Class == "Reptilia" | lpi_data$Class == "Amphibia") & lpi_data$T_realm == "Palearctic"
sum(selected_pops, na.rm = T)
infile_name <- create_infile(lpi_data, index_vector=selected_pops, name="T_Palearctic_herps")

selected_pops = (lpi_data$Class == "Reptilia" | lpi_data$Class == "Amphibia") & lpi_data$T_realm == "Neotropical"
sum(selected_pops, na.rm = T)
infile_name <- create_infile(lpi_data, index_vector=selected_pops, name="T_Neotropical_herps")

selected_pops = (lpi_data$Class == "Reptilia" | lpi_data$Class == "Amphibia") & lpi_data$T_realm == "Nearctic"
sum(selected_pops, na.rm = T)
infile_name <- create_infile(lpi_data, index_vector=selected_pops, name="T_Nearctic_herps")
```
